---
title: "Taylors_Law_Chapter_02"
author: "Soren Maret"
date: "2025.04.02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyr)
library(tidyverse)
library(dplyr)
library(ggplot2)
```

##GOAL##: Do all realivent transformations to our data such as log transformations and sort out all genes that didn't change or returned a 0.0000 for their mean. 
```{r Loading in Data}
load("/scratch/Shares/rinnclass/MASTER_CLASS/STUDENTS/genehomies/TPL_ANALYSIS/RData/Chapter_01.RData")
##our envionrment should be identical to where we left of in chapter_01
```

Great! recall how in our plots that we generated in chapter 01 were super dense close to zero and thinned out as we went away from the x-axis? This indicates that few genes are very highly expressed. precisely, if there were 10 genes in our data, then 50% of our expression would be from one gene. 

To get around this we need to log transform our data using the natural logarithm to get our slope and intercept values described in chapter 01.
```{r log transformations}
#we will create value lists that we will merge into our taylor_values dataframes later
#in R the log function is the natural log, how lovely!
#heres the log_mean:
log_m_human <- log(means_human)
#heres the log_sd:
log_sd_human <- log(sd_human)

#Now we can just repeate this for mice
#heres the log_mean:
log_m_mouse <- log(means_mouse)
#heres the log_sd:
log_sd_mouse <- log(sd_mouse)
```

Great, we could do the gut check plotting we did earlier or we could just read eveything into our taylot_values dataframes just to make life easy
```{r adding our log value lists back into our dataframes}
#for humans
taylor_values_human <- taylor_values_human %>%
  mutate(log_m = log_m_human)
taylor_values_human <- taylor_values_human %>%
  mutate(log_sd = log_sd_human)
#for mice
taylor_values_mouse <- taylor_values_mouse %>%
  mutate(log_m = log_m_mouse)
taylor_values_mouse <- taylor_values_mouse %>%
  mutate(log_sd = log_sd_mouse)
```

Note how some rows have returned a inf or -inf value for their results. This is fine as ggplot just ignore these, but we should remove it.
To do this we will just remove all rows in our mean or standard deviation collums that are zero. While we are at it lets rename our dataframes after the sorting.
```{r sorting out all inf and 0.0000 values from our data}
#humans first!
taylor_values_human_02 <- taylor_values_human[taylor_values_human$mean != 0, ]
taylor_values_human_02 <- taylor_values_human[taylor_values_human$sd != 0, ]
#now mice!
taylor_values_mouse_02 <- taylor_values_mouse[taylor_values_mouse$mean != 0, ]
taylor_values_mouse_02 <- taylor_values_mouse[taylor_values_mouse$sd != 0, ]
```

That got rid of a bunch of our data. now we can run our regressions. For this we will use ggplot2 in chapter_03!
```{r}
save(taylor_values_human_02, taylor_values_mouse_02, file = "/scratch/Shares/rinnclass/MASTER_CLASS/STUDENTS/genehomies/TPL_ANALYSIS/RData/Chapter_02.RData")
```

