---
output: html_document
editor_options: 
  chunk_output_type: console

Date: "2025_03_19"
output: html_document
editor_options: 
  chunk_output_type: console

output: html_document
editor_options: 
  chunk_output_type: console
---
## Title: "Genehomies Final project"
## Authors: 
- [ ] Soren Maret 
- [ ] Kienan Higgins
- [ ] ReAnna Owens
- [ ] Karina Li
- [ ] Caleb Maddry
- [ ] - [Surina Gidha](https://github.com/Surina14)
---

#These are the packages and knit commands we need

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, echo = TRUE)
library(IRanges)
library(dplyr)
library(tidyr)
library(tibble)
library(readr)
library(ggplot2)
library(purrr)
library(magrittr)
library(pheatmap)
library(textshape)
library(Rcpp)
library(DESeq2)
library(matrixStats)
```

# Importing Counts Data

```{r import counts from Salmon}
## Loading Salmon counts
counts_matrix <- read.table("DATA/MOUSE/salmon.merged.gene_counts.tsv", header=TRUE, row.names=1)

## creating a data frame named g2s
g2s <- data.frame(
  gene_id = rownames(counts_matrix),    ### Extracts row names in this case gene IDs
  gene_name = counts_matrix[, 1]        ### Extracts the gene names
)

## Removing gene_name column for analyses
counts_matrix <- counts_matrix[ , -1]

## Rounds the values in counts_matrix to the nearest integer
counts_integer <- round(counts_matrix)
```

# Creating a sample sheet for deseq from a counts file

```{r creating deseq sample sheet (coldata)}
## Setting up data frame with cols from counts matrix
deseq_samples <- data.frame(
  sample_id = colnames(counts_matrix))

## Splitting the sample names (sample_id) into components
split_values <- strsplit(deseq_samples$sample_id, "_")

## This extracts specific parts of each sample name from split_values, which was created earlier using strsplit(deseq_samples$sample_id, "_").
## The sapply() function applies the provided function to each element of split_values, which is a list.
time_values <- sapply(split_values, function(x) x[[2]])
replicate_values <- sapply(split_values, function(x) x[[3]])

## This adds two new columns (time_point and replicate) to the existing deseq_samples data frame.
## These columns store experiment metadata, which is essential for DESeq2 analysis.
deseq_samples$time_point <- time_values
deseq_samples$replicate <- replicate_values

## Converting time_point and replicate into factors.
## In R, a factor is a categorical variable, which is essential for DESeq2 to properly interpret the experimental conditions.
deseq_samples$time_point <- factor(deseq_samples$time_point)
deseq_samples$replicate <- factor(deseq_samples$replicate)
```

# Running DESEQ2

```{r running deseq2}
## # testing sample sheet and counts are arranged properly 
stopifnot(all(colnames(counts_integer) == rownames(deseq_samples$sample_id)))

## This line creates a DESeq2 dataset (dds_time_point) from the count matrix and metadata, which is a crucial step before performing differential expression analysis.
dds_time_point <- DESeqDataSetFromMatrix(countData = counts_integer,
                              colData = deseq_samples,
                              design = ~ time_point)

## Run time point model
dds_time_point <- DESeq(dds_time_point)
```

# Extracting results for each time point from DESEQ dds

```{r compile DESEQ2 results}
## setting up result names object for input to forloop
result_names <- resultsNames(dds_time_point)
results_names <- result_names[-1]

## Setting up a dataframe
res_df <- data.frame("gene_id" = character(), 
                     "baseMean" = numeric(), 
                     "log2FoldChange" = numeric(), 
                     "lfcSE" = numeric(),
                     "stat" = numeric(),
                     "pvalue" = numeric(),
                     "padj" = numeric(),
                     "gene_name" = character(),
                     "result_name" = character())

## run for loop to retreive results from all time comparisons
for(i in 1:length(results_names)) {
  results_name <- results_names[i]
  res <- results(dds_time_point, name = results_name)
  tmp_res_df <- res %>% 
    as.data.frame() %>%
    rownames_to_column("gene_id") %>%
    merge(g2s) %>%
    mutate(result_name = results_name)
  res_df <- bind_rows(res_df, tmp_res_df)
  
}
```

# Here we are filtering to significant results P <0.05 and LFC abs(1)

```{r filtering DESEQ results to LFC > abs(1)& P <0.05}

filtered_res_df <- res_df %>%
  filter(padj < 0.05, abs(log2FoldChange) > 1)

unique_sig_genes <- unique(filtered_res_df$gene_id)
```

# Import TPM file from Salmon

```{r import counts and TPM from Salmon}
## load and prepare TPM file from Salmon
TPM <- read.table("DATA/MOUSE/salmon.merged.gene_tpm.tsv", header=TRUE, row.names=1)

## Removes the gene_name column from TPM
TPM <- TPM %>%
  select(-gene_name)

## Filtering for TPM of at least 1 across samples
TPM_filtered <- TPM[rowSums(TPM) > 1, ]

any(rowSums(TPM_filtered) < 1)
any(rowSums(TPM) < 1)

```


